\pkg{sdmTMB}: An \proglang{R} Package for Fast, Flexible, and User-Friendly Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields

### 2023-10-20 doc/paper-jss/sdmTMB-paper.Rnw
Fix #15
delta method
Cleanup

% Details are beyond the scope of this paper and are not necessary to use the software discussed here, but the idea is that the solution to a specific SPDE is a GRF with a Matérn covariance function and this `trick' enables one to efficiently fit approximations to GRFs to large spatial datasets \citep{lindgren2011}.

@@

% We provide a more detailed comparison of related software packages in Table~\ref{tab:functionality} and the Discussion (Section~\ref{discussion}).

@@

% TODO: there is now some duplication between these two paragraphs!?

@@

% Assuming $\nu = 1$, the parameters $\kappa$ and $\tau$ can be converted to two useful values: correlation range ($\sqrt{8} / \kappa$) and marginal standard deviation The range is the distance that correlation decays to $\approx 0.13$ .

@@

Using $p$ to denote the probability of a non-zero observation and $r$ to represent the expected rate for positive data, we can construct two linear predictors ($\eta_1$ and $\eta_2$) in link space
\[
\begin{aligned}
% \operatorname{logit} (p) &= \boldsymbol{X_1} \boldsymbol{\beta_1} + \ldots,\\
% \log (r) &= \boldsymbol{X_2} \boldsymbol{\beta_2} + \ldots.
\operatorname{logit} (p) &= \eta_1,\\
\log (r) &= \eta_2.
\end{aligned}
\]
We can relate $p$ and $r$ to the data via a Bernoulli and positive likelihood distribution (e.g., lognormal or gamma with $\phi$ representing a generic dispersion parameter).
We use $\mathrm{I}(.)$ to denote an indicator function, which is $1$ if $y > 0$ and 0 if $y = 0$.
\[
\begin{aligned}
\mathrm{I}(y > 0) &\sim \operatorname{Bernoulli} \left( p \right),\\
y | y > 0 &\sim \operatorname{Positive-distribution} \left( r, \phi \right).
\end{aligned}
\]

@@

An additional delta-model is possible that has several advantages over the logit-log delta model \citep{thorson2018poisson}.
The primary advantage is that both linear predictors use a log link so the linear predictors can be added in link space and the partial effect of a coefficient from both linear predictors can be combined.
In these models, the linear predictors represent a theoretical group number $n$ and a theoretical weight (e.g., mass) per group $w$ \citep{thorson2018poisson},
\[
\begin{aligned}
% \log (n) &= \boldsymbol{X_1} \boldsymbol{\beta_1} + \ldots,\\
% \log (w) &= \boldsymbol{X_2} \boldsymbol{\beta_2} + \ldots.
\log (n) &= \eta_1,\\
\log (w) &= \eta_2.
\end{aligned}
\]
Note that the first linear predictor has a log link, \emph{not} a logit link, and the linear predictor predicts group number $n$, \emph{not} positive observation probability $p$.
These get transformed \citep{thorson2018poisson} via
\[
\begin{aligned}
p &= 1 - \exp(- \exp( \log n)),\\
r &= \frac{n w}{p}.
\end{aligned}
\]
The first part is the complementary-log-log inverse link \citep[][p.~31]{mccullagh1989} (which has its roots in a Poisson process) but the group number $n$ also enters into the expected positive rate $r$.
These probabilities $p$ and positive rates $r$ are then entered into a Bernoulli and positive observation likelihood as before
\[
\begin{aligned}
\mathrm{I}(y > 0) &\sim \operatorname{Bernoulli} \left( p \right),\\
y | y > 0 &\sim \operatorname{Positive-distribution} \left( r, \phi \right).
\end{aligned}
\]

% \sdmTMB\ fits GLMMs to spatial or spatiotemporal geostatistical data.
% Geostatistical data refers to data observed at specific spatial coordinates
% reflecting some underlying spatial process \citep{rossi1992, diggle2007}. These
% data can be collected across discrete points in time. The process component of
% an \sdmTMB\ model can be formed by any combination of main (``fixed'')
% effects \(\bm{\beta}\), an offset \(O_{\bm{s},t}\), IID (independent and
% identically distributed) random intercepts \(\alpha_g\), spatial intercept
% random fields \(\omega_{\bm{s}}\), spatiotemporal intercept random fields
% \(\epsilon_{\bm{s},t}\), time-varying effects \(\bm{\gamma_t}\), and spatially
% varying effects \(\zeta_{\bm{s}}\) (Figure~\ref{fig:diagram}). These components
% are discussed in more detail via our examples below.

% This process component is combined with an observation error family (e.g.,
% Gaussian, gamma, binomial, Tweedie) and link function \(g\) (e.g., identity,
% log, logit) as in any generalized linear model (GLM) \citep{mccullagh1989}. Some
% families can be combined into a two-part ``delta'' or ``hurdle'' model
% \citep{aitchison1955} to model the zero vs.~non-zero observations separately
% from the positive observations. See the model description vignette for an
% expanded description of the statistical model
% (\code{browseVignettes("sdmTMB")} or see the
% \href{https://pbs-assess.github.io/sdmTMB/articles/}{rendered version}).

% By default, if spatiotemporal fields are included, they are assumed IID; however, additional options allow them to be modelled as a random walk or first-order autoregressive, AR(1), process (Figure~\ref{fig:diagram}).
% Turning off both spatial and spatiotemporal effects allows for comparison with standard non-spatial GLMs or GLMMs.

@@

% TODO: () on end of functions!? see flowchart vs. rest
% TODO: blah.sdmTMB() for S3 methods?

@@

@@

% Describe modular design philosopy

% The design goal of...

% flexible, modular, familiar, make SPDE accessible and fast

% combine functionality commonly needed together

% The \sdmTMB\ model is fit by maximum marginal likelihood. Internally,
% a \TMB\ \citep{kristensen2016} model template is used to calculate the
% marginal log likelihood and its gradient, and the negative log likelihood is
% minimized via the non-linear optimization routine \code{stats::nlminb()} in
% \R\ \citep{gay1990, r2021}. Random effects are estimated at values that
% maximize the log likelihood conditional on the estimated fixed effects and are
% integrated over via the Laplace approximation \citep{kristensen2016}. After
% rapid model exploration with maximum likelihood, one can optionally pass an
% \sdmTMB\ model to the \R\ package \pkg{rstan} \citep{carpenter2017,
% rstan} via \pkg{tmbstan} \citep{monnahan2018} to estimate the joint posterior
% distribution for Bayesian inference.

@@

% TODO rename this? move this? add additional sections above!?

@@

% \subsection{Software dependencies}\label{dependencies}

% \sdmTMB\ relies on several well-established \R\ packages (Figure~\ref{fig:flowchart}).
% \sdmTMB\ uses \pkg{fmesher} \citep{fmesher} and \INLA\ \citep{lindgren2015} to generate the triangulation mesh and input matrices for the SPDE calculations.
% The SPDE precision matrix calculations themselves are performed by functions within \TMB\ \citep{kristensen2016} (\code{R\_inla} namespace).
% Data structures required to implement penalized smoothers are constructed with \fct{smooth2random} from \pkg{mgcv} \citep{wood2017a}, similarly to \pkg{gamm4} \citep{wood2020} and \pkg{brms} \citep{brms}.
% The statistical model itself is written as a \proglang{C++} template for \TMB.
% \TMB\ returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016}.
% The negative marginal log likelihood is minimized via the non-linear optimization routine \fct{stats::nlminb} \citep{gay1990} in \R\ with additional optimization carried out via a Newton optimization step to find the Hessian using \fct{stats::optimHess} \citep{gay1990, r2021}.
% After rapid model exploration with maximum likelihood, one can optionally pass an \sdmTMB\ model to the \R\ package \pkg{rstan} \citep{carpenter2017, rstan} via \pkg{tmbstan} \citep{monnahan2018} to samples from the joint posterior distribution for Bayesian inference.
% % TODO: visreg + ggeffects effects, etc.?

@@

We can also specify additional arguments, in this case passed to \fct{fmesher::fm\_mesh\_2d\_inla}: a maximum triangle edge (\code{max.edge}) length of 10 km and 40 km for the inner and outer mesh, and an \code{offset} width of 10 km and 40 km for the inner and outer meshes borders.
% Alternatively, we could have created any mesh via \INLA\ or \pkg{fmesher} and supplied it to the \code{mesh} argument in \fct{make\_mesh}.
For more irregularly shaped areas, we could have specified a non-convex hull with \code{convex} and \code{concave} arguments.
The triangle edge length should be several times smaller than the range and the outer boundary should extend at least as far as the range (REF).

@@

We can test for spatial autocorrelation with a Moran's I test or with a visual inspection of the residuals calculated from the \fct{residuals} method.
We omit example code for Moran's I calculation for brevity but include it in the code supplement.
% TODO: reword this?
We see strong evidence for spatial autocorrelation for the model without a random field (p $<$ 0.01) but no evidence for the model with a random field (p = 0.84) suggesting that residual spatial autocorrelation is alleviated by including the random field.

@@

We can explore four alternative families that may better represent the data.
Each alternative family uses a delta model formulation as described in Section \ref{sec:delta}.
% The Poisson-link models depart from conventional delta models in treating the probability of occurrence at a given location ($p_i$) as a Poisson process rather than a Bernoulli one, $p_i = 1 - \exp(a_i \cdot n_i)$, where $n_i$ is the total number of individuals available to be encountered, and $a_i$ is a measure of sampling effort \citep{thorson2018poisson}.
Note that we choose to omit the spatiotemporal random fields from the binomial model component (\code{spatiotemporal = list("off", "iid")}) because initial fits indicated this random field was collapsing to zero indicating minimal changes in encounter spatial distribution from year to year.

% TODO EQUATIONS NEED TO INTRO THE POISSON LINK AS BRIEFLY AS POSSIBLE. SEE sdmTMB VIGNETTE.

@@

% TODO: Need description of anisotropy and why it's important here or somewhere else

@@

% TODO Propose moving this to a short paragraph `Prediction` to highlight modularity

@@

% TODO component confusing:

@@

We can generate an area-weighted population index (e.g., a relative or absolute index of abundance or biomass) that is independent of sampling locations by predicting from the model on a grid covering the area of interest and summing the predicted biomass with the \fct{get\_index} function (Figure \ref{fig:dog-index-plot}).
We supply the grid cell area (4 km\textsuperscript{2}) to the \code{area} argument and specify \code{bias\_correct = TRUE} to enable a bias correction needed due to the non-linear transformation of the random effects \citep{thorson2016bias}.
% TODO and the Laplace?

@@

Our data are read in from supplementary data and contain columns for spatial coordinates (in an Albers projection for North America and divided by 100000 to give units of 100 km), year, year as a factor, and the annual NAO value:
<<owl-data>>=
snow <- readRDS(here::here("data/SNOW_data.rds"))
@

@@

There are many \R\ packages capable of fitting geostatistical spatial or spatiotemporal models \citep[e.g.,][]{heaton2019}.
\sdmTMB, \pkg{VAST}, \INLA/\pkg{inlabru}, and \pkg{spaMM} \citep{rousset2014} are the most closely related, as they all provide a user interface to SPDE-based GMRF models.
In our software comparison (Table \ref{tab:functionality}), we also include \pkg{mgcv} as it can be adapted to use the SPDE \citep{miller2019} and \pkg{spBayes} \citep{finley2007, finley2015} since it is a prominent package that can fit related predictive-process models without the SPDE.
\sdmTMB, \pkg{VAST}, and \pkg{mgcv} can estimate anisotropic covariance whereas \INLA/\pkg{inlabru} and \pkg{spBayes} are currently limited to isotropic covariance.
\sdmTMB\ and \pkg{mgcv} focus on univariate response data, whereas \pkg{VAST}, \INLA/\pkg{inlabru}, \pkg{spaMM}, and \pkg{spBayes} extend to multivariate responses with various limitations.
To our knowledge, \pkg{VAST} is the only package to implement spatial \citep{thorson2015} and spatial dynamic factor analysis \citep{thorson2016} and spatial empirical orthogonal function (EOF) regression \citep{thorson2020}.
Of these packages, only \sdmTMB\ and \pkg{inlabru} can currently fit threshold (e.g., hockey-stick) covariate relationships.
\pkg{spaMM} is limited to a spatial random field (i.e., does not fit spatiotemporal fields) and \pkg{spBayes} implements spatiotemporal fields, but only as a random walk.
There is considerable variability in the available observation likelihoods across packages (Table \ref{tab:functionality}).
% We provide comparisons of the syntax and the reproducibility of results from models fit using \sdmTMB, \INLA, and \pkg{inlabru} in Appendix~\ref{app:inla}.

@@

\centering
\fontsize{8}{10}\selectfont
\begin{threeparttable}
\begin{tabular}[t]{lllllll}
\toprule
 & \sdmTMB\ & \pkg{VAST} & \INLA/\pkg{inlabru} & \pkg{mgcv} & \pkg{spBayes} & \pkg{spaMM}\\
\midrule
Time-varying coefficients & \checkmark & --$^1$ & \checkmark & \checkmark & \checkmark & --\\
Spatially varying coefficients (SVC) & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & --\\
GAMs$^2$ & \checkmark & -- & \checkmark & \checkmark & -- & --\\
Threshold covariates & \checkmark & -- & \checkmark$^3$ & -- & -- & --\\
Offsets & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark\\
\addlinespace
Spatiotemporal fields & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark$^4$ & --\\
Spatial + spatiotemporal fields & \checkmark & \checkmark & \checkmark & \checkmark & -- & --\\
Anisotropy & \checkmark & \checkmark & -- & \checkmark & -- & --\\
Correlation barriers & \checkmark & \checkmark & \checkmark & \checkmark & -- & --\\
Separate range parameters for fields & \checkmark & -- & \checkmark & \checkmark & -- & --\\
Share range parameters across fields & \checkmark & \checkmark & \checkmark & -- & -- & --\\
SPDE-based & \checkmark & \checkmark & \checkmark & \checkmark$^5$ & --$^6$ & \checkmark\\
\addlinespace
NB1 distribution & \checkmark & -- & \checkmark & \checkmark & -- & \checkmark\\
NB2 distribution & \checkmark & \checkmark$^7$ & \checkmark & \checkmark & -- & \checkmark\\
Zero-truncated distributions & \checkmark & -- & \checkmark & -- & -- & \checkmark\\
Zero-inflated distributions & \checkmark & \checkmark & \checkmark & -- & -- & \checkmark\\
Tweedie distribution & \checkmark & \checkmark & \checkmark & \checkmark$^8$ & -- & --\\
Student-t distribution & \checkmark & -- & \checkmark & \checkmark & -- & --\\
Censored Poisson distribution & \checkmark & -- & \checkmark & -- & -- & --\\
log Gaussian Cox processes & --$^9$ & --$^9$ & \checkmark & --$^9$ & --$^9$ & --$^9$\\
Multivariate responses & -- & \checkmark & \checkmark & -- & \checkmark & \checkmark\\
Built-in delta/hurdle models & \checkmark & \checkmark & \checkmark & --$^{10}$ & -- & \checkmark\\
Poisson-link delta models & \checkmark & \checkmark & \checkmark & -- & -- & --\\
Likelihood weights & \checkmark & -- & \checkmark & \checkmark & \checkmark & \checkmark\\
Maximum/marginal likelihood & \checkmark & \checkmark & -- & \checkmark & -- & --\\
\addlinespace
Bayesian/optionally Bayesian & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & --\\
Priors/penalties & \checkmark & -- & \checkmark & -- & \checkmark & --\\
Matern PC priors & \checkmark & -- & \checkmark & -- & -- & --\\
\addlinespace
Spatial (or spatial dynamic) factor analysis & -- & \checkmark & -- & -- & -- & --\\
Empirical Orthogonal Function (EOF) analysis & -- & \checkmark & -- & -- & -- & --\\
Built-in area-weighted index standardization & \checkmark & \checkmark & -- & -- & -- & --\\
Built-in cross-validation & \checkmark & -- & -- & -- & -- & --\\
\bottomrule
\end{tabular}
\end{threeparttable}
\label{tab:functionality}

\caption{Comparison of functionality between several \R\ packages that
  can fit geostatistical GLMMs.
Notes:
$^1$Technically possible but non-trivial.
$^2$Penalized smoother GAMs that determine `wiggliness'.
$^3$\pkg{inlabru} but not \INLA.
$^4$Spatiotemporal fields as random walk only.
$^5$SPDE approach as in \citet{miller2019}.
$^6$Does have predictive process knots.
$^7$Zero-inflated NB2 only.
$^8$Tweedie power parameter fixed for \fct{mgcv::gamm}.
$^9$Possible as log-linked Poisson GLMM with aggregated data.
$^{10}$Hurdle models possible by fitting components separately.
}
\end{table}

We ran a simple speed comparison between \sdmTMB, \INLA/\pkg{inlabru}, and \pkg{mgcv} for fitting an SPDE spatial random field model to 1000 data points with Gaussian error across a range of mesh resolutions (Figure~\ref{fig:timing}, Appendix \ref{app:speed}).
In this test, \sdmTMB\ was the fastest across all mesh resolutions; however, \INLA/\pkg{inlabru} and \pkg{spaMM} were less affected by mesh resolution than \sdmTMB.
\pkg{mgcv} was most affected by mesh resolution.
Our test was restricted to one core and default \R\ algebra libraries.
% all packages could run faster with optimized libraries and parallel processing.
% Results with optimized math libraries on one core (\pkg{openBLAS}: \citet{openblas} and \pkg{PARDISO}: \citet{pardiso7.2a}) resulted in a \textasciitilde10\% speed increase for \sdmTMB\ and \pkg{inlabru} and a \textasciitilde7--9-fold speed increase for \pkg{mgcv}.

@@

% (4) continuous-time models \citep[e.g.,][]{blangiardo2015book}, (5) non-Gaussian random fields \citep[e.g.,][]{anderson2019}, and (6) multivariate responses.
% The included \TMB\ \code{.cpp} file is flexible and provides a tested model template that can be modified to add additional features.
% TODO random slopes looks gnarly to do generically? EW: -- agree, I mostly have looked at glmmTMB / brms implementations
% removed random slopes -- another feature that could be added is 3D models, e.g. https://doi.org/10.1016/j.spasta.2023.100750

>>>>

@@

@@

@@

@@

Using $p$ to denote the probability of a non-zero observation and $r$ to represent the expected rate for positive data, we can construct two linear predictors ($\eta_1$ and $\eta_2$) in link space
\[
\begin{aligned}
\operatorname{logit} (p) &= \eta_1,\\
\log (r) &= \eta_2.
\end{aligned}
\]
We can relate $p$ and $r$ to the data via a Bernoulli and positive likelihood distribution (e.g., lognormal or gamma with $\phi$ representing a generic dispersion parameter).
We use $\mathrm{I}(.)$ to denote an indicator function, which is $1$ if $y > 0$ and 0 if $y = 0$.
\[
\begin{aligned}
\mathrm{I}(y > 0) &\sim \operatorname{Bernoulli} \left( p \right),\\
y | y > 0 &\sim \operatorname{Positive-distribution} \left( r, \phi \right).
\end{aligned}
\]

@@

An additional delta-model is possible that has several advantages over the logit-log delta model \citep{thorson2018poisson}.
The primary advantage is that both linear predictors use a log link so the linear predictors can be added in link space and the partial effect of a coefficient from both linear predictors can be combined.
In these models, the linear predictors represent a theoretical group number $n$ and a theoretical weight (e.g., mass) per group $w$ \citep{thorson2018poisson},
\[
\begin{aligned}
\log (n) &= \eta_1,\\
\log (w) &= \eta_2.
\end{aligned}
\]
Note that the first linear predictor has a log link, \emph{not} a logit link, and the linear predictor predicts group number $n$, \emph{not} positive observation probability $p$.
These get transformed \citep{thorson2018poisson} via
\[
\begin{aligned}
p &= 1 - \exp(- \exp( \log n)),\\
r &= \frac{n w}{p}.
\end{aligned}
\]
The first part is the complementary-log-log inverse link \citep[][p.~31]{mccullagh1989} (which has its roots in a Poisson process) but the group number $n$ also enters into the expected positive rate $r$.
These probabilities $p$ and positive rates $r$ are then entered into a Bernoulli and positive observation likelihood as before
\[
\begin{aligned}
\mathrm{I}(y > 0) &\sim \operatorname{Bernoulli} \left( p \right),\\
y | y > 0 &\sim \operatorname{Positive-distribution} \left( r, \phi \right).
\end{aligned}
\]

@@

@@

\textbf{TODO: generalized delta method! and \citep{zheng2021}}

@@

@@

@@

@@

We can also specify additional arguments, in this case passed to \fct{fmesher::fm\_mesh\_2d\_inla}: a maximum triangle edge (\code{max.edge}) length of 10 km and 40 km for the inner and outer mesh, and an \code{offset} width of 10 km and 40 km for the inner and outer meshes borders.
For more irregularly shaped areas, we could have specified a non-convex hull with \code{convex} and \code{concave} arguments.
The triangle edge length should be several times smaller than the range and the outer boundary should extend at least as far as the range (REF).

@@

We can test for spatial autocorrelation with a Moran's I test or with a visual inspection of the residuals calculated from the \fct{residuals} method.
We omit example code for Moran's I calculation for brevity but include it in the code supplement.
We see strong evidence for spatial autocorrelation for the model without a random field (p $<$ 0.01) but no evidence for the model with a random field (p = 0.84) suggesting that residual spatial autocorrelation is alleviated by including the random field.

@@

We can explore four alternative families that may better represent the data.
Each alternative family uses a delta model formulation as described in Section \ref{sec:delta}.
Note that we choose to omit the spatiotemporal random fields from the binomial model component (\code{spatiotemporal = list("off", "iid")}) because initial fits indicated this random field was collapsing to zero indicating minimal changes in encounter spatial distribution from year to year.

@@

@@

@@

@@

We can generate an area-weighted population index (e.g., a relative or absolute index of abundance or biomass) that is independent of sampling locations by predicting from the model on a grid covering the area of interest and summing the predicted biomass with the \fct{get\_index} function (Figure \ref{fig:dog-index-plot}).
We supply the grid cell area (4 km\textsuperscript{2}) to the \code{area} argument and specify \code{bias\_correct = TRUE} to enable a bias correction needed due to the non-linear transformation of the random effects \citep{thorson2016bias}.

@@

Our data are read in from supplementary data and contain columns for spatial coordinates (in an Albers projection for North America and divided by 100000 to give units of 100 km), year, year as a factor, and the annual NAO value:

<<owl-data>>=
snow <- readRDS(here::here("data/SNOW_data.rds"))
@

@@

There are many \R\ packages capable of fitting geostatistical spatial or spatiotemporal models \citep[e.g.,][]{heaton2019}.
\sdmTMB, \pkg{VAST}, \INLA/\pkg{inlabru}, and \pkg{spaMM} \citep{rousset2014} are the most closely related, as they all provide a user interface to SPDE-based GMRF models.
In our software comparison (Table \ref{tab:functionality}), we also include \pkg{mgcv} as it can be adapted to use the SPDE \citep{miller2019} and \pkg{spBayes} \citep{finley2007, finley2015} since it is a prominent package that can fit related predictive-process models without the SPDE.
\sdmTMB, \pkg{VAST}, and \pkg{mgcv} can estimate anisotropic covariance whereas \INLA/\pkg{inlabru} and \pkg{spBayes} are currently limited to isotropic covariance.
\sdmTMB\ and \pkg{mgcv} focus on univariate response data, whereas \pkg{VAST}, \INLA/\pkg{inlabru}, \pkg{spaMM}, and \pkg{spBayes} extend to multivariate responses with various limitations.
To our knowledge, \pkg{VAST} is the only package to implement spatial \citep{thorson2015} and spatial dynamic factor analysis \citep{thorson2016} and spatial empirical orthogonal function (EOF) regression \citep{thorson2020}.
Of these packages, only \sdmTMB\ and \pkg{inlabru} can currently fit threshold (e.g., hockey-stick) covariate relationships.
\pkg{spaMM} is limited to a spatial random field (i.e., does not fit spatiotemporal fields) and \pkg{spBayes} implements spatiotemporal fields, but only as a random walk.
There is considerable variability in the available observation likelihoods across packages (Table \ref{tab:functionality}).

@@

\centering
\fontsize{8}{10}\selectfont
\begin{threeparttable}
\begin{tabular}[t]{lllllll}
\toprule
 & \sdmTMB\ & \pkg{VAST} & \INLA/\pkg{inlabru} & \pkg{mgcv} & \pkg{spBayes} & \pkg{spaMM}\\
\midrule
Time-varying coefficients & \checkmark & --$^1$ & \checkmark & \checkmark & \checkmark & --\\
Spatially varying coefficients (SVC) & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & --\\
GAMs$^2$ & \checkmark & -- & \checkmark & \checkmark & -- & --\\
Threshold covariates & \checkmark & -- & \checkmark$^3$ & -- & -- & --\\
Offsets & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark\\
\addlinespace
Spatiotemporal fields & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark$^4$ & --\\
Spatial + spatiotemporal fields & \checkmark & \checkmark & \checkmark & \checkmark & -- & --\\
Anisotropy & \checkmark & \checkmark & -- & \checkmark & -- & --\\
Correlation barriers & \checkmark & \checkmark & \checkmark & \checkmark & -- & --\\
Separate range parameters for fields & \checkmark & -- & \checkmark & \checkmark & -- & --\\
Share range parameters across fields & \checkmark & \checkmark & \checkmark & -- & -- & --\\
SPDE-based & \checkmark & \checkmark & \checkmark & \checkmark$^5$ & --$^6$ & \checkmark\\
\addlinespace
NB1 distribution & \checkmark & -- & \checkmark & \checkmark & -- & \checkmark\\
NB2 distribution & \checkmark & \checkmark$^7$ & \checkmark & \checkmark & -- & \checkmark\\
Zero-truncated distributions & \checkmark & -- & \checkmark & -- & -- & \checkmark\\
Zero-inflated distributions & \checkmark & \checkmark & \checkmark & -- & -- & \checkmark\\
Tweedie distribution & \checkmark & \checkmark & \checkmark & \checkmark$^8$ & -- & --\\
Student-t distribution & \checkmark & -- & \checkmark & \checkmark & -- & --\\
Censored Poisson distribution & \checkmark & -- & \checkmark & -- & -- & --\\
log Gaussian Cox processes & --$^9$ & --$^9$ & \checkmark & --$^9$ & --$^9$ & --$^9$\\
Multivariate responses & -- & \checkmark & \checkmark & -- & \checkmark & \checkmark\\
Built-in delta/hurdle models & \checkmark & \checkmark & \checkmark & --$^{10}$ & -- & \checkmark\\
Poisson-link delta models & \checkmark & \checkmark & \checkmark & -- & -- & --\\
Likelihood weights & \checkmark & -- & \checkmark & \checkmark & \checkmark & \checkmark\\
Maximum/marginal likelihood & \checkmark & \checkmark & -- & \checkmark & -- & --\\
\addlinespace
Bayesian/optionally Bayesian & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & --\\
Priors/penalties & \checkmark & -- & \checkmark & -- & \checkmark & --\\
Matern PC priors & \checkmark & -- & \checkmark & -- & -- & --\\
\addlinespace
Spatial (or spatial dynamic) factor analysis & -- & \checkmark & -- & -- & -- & --\\
Empirical Orthogonal Function (EOF) analysis & -- & \checkmark & -- & -- & -- & --\\
Built-in area-weighted index standardization & \checkmark & \checkmark & -- & -- & -- & --\\
Built-in cross-validation & \checkmark & -- & -- & -- & -- & --\\
\bottomrule
\end{tabular}
\end{threeparttable}

\caption{Comparison of functionality between several \R\ packages that
  can fit geostatistical GLMMs.
Notes:
$^1$Technically possible but non-trivial.
$^2$Penalized smoother GAMs that determine `wiggliness'.
$^3$\pkg{inlabru} but not \INLA.
$^4$Spatiotemporal fields as random walk only.
$^5$SPDE approach as in \citet{miller2019}.
$^6$Does have predictive process knots.
$^7$Zero-inflated NB2 only.
$^8$Tweedie power parameter fixed for \fct{mgcv::gamm}.
$^9$Possible as log-linked Poisson GLMM with aggregated data.
$^{10}$Hurdle models possible by fitting components separately.
}
\label{tab:functionality}
\end{table}

We ran a simple speed comparison between \sdmTMB, \INLA/\pkg{inlabru}, and \pkg{mgcv} for fitting an SPDE spatial random field model to 1000 data points with Gaussian error across a range of mesh resolutions (Figure~\ref{fig:timing}, Appendix \ref{app:speed}).
In this test, \sdmTMB\ was the fastest across all mesh resolutions; however, \INLA/\pkg{inlabru} and \pkg{spaMM} were less affected by mesh resolution than \sdmTMB.
\pkg{mgcv} was most affected by mesh resolution.
Our test was restricted to one core and default \R\ algebra libraries.

@@

% another feature that could be added is 3D models, e.g. https://doi.org/10.1016/j.spasta.2023.100750
